---
title:  "fctbases vignette"
author: "Niels Olsen"
date: "3/12/2020"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{fctbases vignette}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Easy-to-use, fast implementations of functional bases for use in functional data analysis and elsewhere.

`fctbases` is a package for R, which implements some of the common linear functional bases such as B-splines and Fourier bases and stores these internally as C++ objects, accesssed from R as normal functions. 
In this way there is no need for initializing an R object every time a basis is used in R. One simply initializes the desired basis, which is returned as an R function that one calls with desired time/evaluation points and possibly coefficients. All calculations are implemented in C++. 
By moving some of computations to the time when objects are initialized, this speeds up some of the computations the even more.
The package takes care of the internal bookkeeping of C++ objects and ensures the validity of these. 

First and second derivatives are also provided using the mathematical formulae for this. This is precise and uses no approximations.  

## Initializing a functional basis
Initialize a basis function by calling an appropiate initialization function, e.g.

`knots <- 0:10 / 10`

`f <- make.bspline.basis(knots, order = 4)`

will return a bspline of order 4 (standard) with equidistant knots from 0 to 1.

`endpoints <- c(0, 1)`

`f <-  make.fourier.basis(endpoints, 10)`

will return a Fourier basis with harmonics up to order 10 (that is, 21 degress of freedom) anchored in 0 and 1. 

Please see the help pages of the different functions for details. 

## Using a fctbasis function
After having generated a fctbasis object, it will return a function like this:

````
function (t, x, deriv = FALSE) 
{
    if (missing(x)) {
        if (deriv > 1L) 
            cpp_eval_D2(basis, t)
        else if (deriv) 
            cpp_eval_D(basis, t)
        else cpp_eval_0(basis, t)
    }
    else {
        if (deriv > 1L) 
            cpp_eval_D2_coefs(basis, t, x)
        else if (deriv) 
            cpp_eval_Dcoefs(basis, t, x)
        else cpp_eval_coefs(basis, t, x)
    }
}
<bytecode: 0x...>
<environment: 0x...>
attr(,"class")
[1] "fctbasis"
````

We see that this function (call it `f`) takes three arguments: 
`t` is a vector of evaluation points, 
`x` are optional coefficients to be multiplied, 
and `deriv` is whether the derivative (wrt. `t`) should be evaluated or not (defaults to false).


`f(t)`: Returns a matrix of the basis function evaluted at time points `t`.

`f(t, x)`: Returns a vector of the basis function evaluted at time points `t`, multiplied by coefficients `x`. Equal to `f(t) %*% x`

`f(t, deriv = T)`: Returns first derivative, $d/dt$ `f(t)`.

`f(t, x, deriv = T)`: Returns first derivative,  $d/dt$ `f(t) %*% x`.

`f(t, deriv = 2)`: Returns second derivative, ${d^2}/ {dt^2}$ `f(t)`.

`f(t, x, deriv = 2)`: Returns second derivative, $d^2 / dt^2$ `f(t) %*% x`.




## Speed
A small code example with speed comparison
```{r pakker, warning=FALSE}
## Libraries
library(fctbases)
library(microbenchmark)
library(splines)
```
```{r speed, echo=TRUE}


## Order 4 B-spline, 13 basis functions (10 intervals + 3)
knots <- 0:10 / 10

## the default way using splines::bs
bsb <- function(x) bs(x, knots = knots[2:10], Boundary.knots = c(knots[1], knots[11]), intercept = T)
## fctbases;
bf <- make.bspline.basis(knots = knots)

## some random coefficients and evaluation points
set.seed(3457)
coefs <- rnorm(13)
y <- sort(runif(100))


bf(0.3457)
bsb(0.3457)

microbenchmark(bf(y), bsb(y), times = 1000)
microbenchmark(bf(y, coefs), bsb(y) %*% coefs, times = 1000)


```


## Installation
The package is available from CRAN: `install.package("fctbases")` or GitHub: "naolsen/fctbases". 
The version on Github may be newer: download the source package or use devtools, e.g. `devtools::install_github("naolsen/fctbases")`. A C++ compiler is required to compile the source.

## Issues
It is currently not possible to save `fctbases` objects as .RData objects (and likely will not be).
Using a `fctbasis` object from a previous session will return an error. 


